---
title: "Rpadrino: an R package to access and use PADRINO, an open access database of Integral Projection Models"
output:  
  word_document:
    toc: no
    keep_md: true
fig_caption: yes
---

A manuscript in preparation for submission to Methods in Ecology and Evolution

Sam C. Levin $^{\ast1,2,3}$, Sanne Evers $^{1,2}$, Tomos Potter$^{3}$, Mayra Pe$\mathrm{\tilde{n}}$a Guerrero $^{1,2}$, Dylan Z. Childs $^4$, Aldo Compagnoni $^{1,2,3}$, Tiffany M. Knight $^{1,2,5\ddagger}$, Roberto Salguero-Gómez $^{3\ddagger}$

$^1$Institute of Biology, Martin Luther University Halle-Wittenberg, Am Kirchtor 1, 06108 Halle (Saale), Germany

$^2$German Centre for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Deutscher Platz 5e, 04103 Leipzig, Germany

$^3$Department of Zoology, 11a Mansfield Rd, University of Oxford, Oxford, OX1 3SZ, UK

$^4$Department of Animal and Plant Sciences, University of Sheffield, Sheffield, S10 2TN, UK

$^5$Department of Community Ecology, Helmholtz Centre for Environmental Research-UFZ, Theodor-Lieser-Straße 4, 06120, Halle (Saale), Germany

$^\ddagger$ Joint senior authors

$^*$Corresponding Author:

Sam C. Levin

Puschstrasse 4, 04103 Leipzig, Germany

email: <levisc8@gmail.com>


```{r echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(purrr)
library(grid)
library(patchwork)
library(ggplot2)
library(Rpadrino)
library(knitr)
library(sf)
library(png)


ipm_pubs <- read.csv('../../metadata/all_ipm_papers_20210806.csv',
                     stringsAsFactors = FALSE) %>% 
  filter(!is.na(Kingdom) & Kingdom != "")

names(ipm_pubs) <- gsub("\\.\\.", "_", names(ipm_pubs))
names(ipm_pubs) <- gsub("\\.$", "", names(ipm_pubs))
names(ipm_pubs) <- gsub("\\.", "_", names(ipm_pubs))

ipm_id <- pmap_chr(.l = data.frame(a = ipm_pubs$Authors,
                                   b = ipm_pubs$Journal, 
                                   c = ipm_pubs$Year),
               .f = function(a, b, c) paste(a, b, c, sep = "_")) 

ipm_pubs <- cbind(ipm_pubs, ipm_id)
ipm_pubs$Year <- as.integer(ipm_pubs$Year)

pub_tot  <- length(unique(ipm_id))

kingdoms <- ipm_pubs %>% 
  group_by(Kingdom) %>%
  summarise(N     = length(unique(Full_citation_APA)),
            n_spp = length(unique(Species))) 


pdb <- pdb_download(save = FALSE)


tab <- pdb$Metadata %>% 
  group_by(kingdom) %>%
  summarise(
    IPMs = length(unique(ipm_id)),
    spp  = length(unique(species_accepted)),
    pap  = length(unique(apa_citation))
  )

tot <- pdb$Metadata %>% 
  summarise(
    kingdom = "Totals",
    IPMs = length(unique(ipm_id)),
    spp  = length(unique(species_accepted)),
    pap  = length(unique(apa_citation))
  ) %>%
  rbind(tab)

n_animalia <- tot[tot$kingdom == "Animalia", "pap"]
n_plantae  <- tot[tot$kingdom == "Plantae", "pap"]
n_tot      <- tot[tot$kingdom == "Totals", "IPMs"]
n_pap      <- tot[tot$kingdom == "Totals", "pap"]

```

# Abstract

1. Discrete time structured population projection models are an important tool for studying population dynamics. Within this field, Integral Projection Models (IPMs) have become a popular method for studying populations structured by continuously distributed traits (_e.g._ height, weight). Databases of discrete time, discrete state structured population models, for example DATLife (life tables) and COMPADRE & COMADRE (matrix population models), have made quantitative syntheses straightforward to implement. These efforts allow researchers to address questions in both basic and applied ecology and evolutionary biology. Since their introduction in 2000, over `r round(sum(pub_tot), -1)` works containing IPMs have been published, offering opportunities for ecological synthesis too. We describe a novel framework to quickly reconstruct these models for subsequent analyses using _Rpadrino_ _R_ package, which serves as an interface to PADRINO, a new database of structured population models. 

2. We introduce an R package, _Rpadrino_, which enables users to download, subset, reconstruct, and extend published IPMs. _Rpadrino_ makes use of recently created software, _ipmr_, to provide an engine to reconstruct a wide array of IPMs from their symbolic representations and conduct subsequent analyses. _Rpadrino_ and _ipmr_ are extensively documented to help users learn their usage. 

3. _Rpadrino_ currently enables users to reconstruct `r n_tot` IPMs from `r n_pap` publications that describe the demography of `r n_animalia` animal and `r n_plantae` plant species. All of these IPMs are tested to ensure they reproduce published estimates. _Rpadrino_ provides an interface to augment PADRINO with external data and modify parameter values, creating a platform to extend models beyond their original purpose while retaining full reproducibility.

4. PADRINO and _Rpadrino_ provide a toolbox for asking new questions and conducting syntheses with peer-reviewed published IPMs. _Rpadrino_ provides a user-friendly interface so researchers do not need to worry about the database structure or syntax, and can focus on their research questions and analyses. Additionally, _Rpadrino_ is thoroughly documented, and provides numerous examples of how to perform analyses which are not included in the package's functionality. 

**Keywords**:  database, demography, elasticity, life history, open access, population dynamics, sensitivity

# Introduction

Demography provides an excellent approach to examine the ecology (Crone et al. 2011), evolutionary biology (Metcalf & Pavard 2007), and conservation biology of any species (Doak & Morris 2001). Environmental conditions and biotic interactions influence vital rates (_e.g._ survival, development, reproduction) across the entire life cycle, which then govern its short-term and long-term performance (Caswell 2001). A variety of methods exist for summarizing vital rates into demographic models; discrete-time, structured population models are among the most popular (Crone et al. 2011, Caswell 2001). Indeed, there is a rich history of using such structured population models across a variety of sub-disciplines in ecology (_e.g._ Leslie 1945, Caswell 2001, Easterling et al. 2000, Adler et al. 2010, Ellner et al. 2016).

In ecology, matrix projection models (MPMs) are the most widely used structured population model. MPMs divide the population into discrete classes corresponding to some trait value (_e.g._ developmental state, age, or size), and then model the population using vital rates computed for each class. Researchers have also recognized that, for some species, vital rates are best predicted as a function of one or more continuous traits (_e.g._ size, height, mass), rather than as a function of discrete classes (Easterling et al. 2000).  Integral projection models (IPMs), which are continuously structured population models, have become an increasingly important tool for ecologists interested in addressing broad biological questions through a demographic lens (Gonzalez et al. 2021). IPMs combine vital rate functions of continuous traits into projection kernels, which describe how the abundance and distribution of trait values in a population change in discrete time (Easterling et al. 2000). IPMs have been used to investigate a variety of topics, such as invasive species spread (_e.g._ Jongejans et al. 2011, Erickson et al. 2017), evolutionary stable strategies (_e.g._ Childs et al. 2004), the effect of climate drivers on population persistence (Salguero-Gómez et al. 2012, Compagnoni et al. 2021a), and linking evolutionary feedbacks to population dynamics (Coulson et al. 2011). 

In order reconstruct and use an IPM, researchers need, at a minimum, the symbolic representation of the model and the associated parameter values. Existing demographic databases enter the transition values directly, rather than a symbolic version of the model and the values associated with the symbols separately. For example, COMPADRE and COMADRE store transition matrices as numeric matrices (sub-matrices corresponding to survival and development ($U$), sexual reproduction ($F$), and asexual reproduction ($C$), and their sum $A$), rather than symbolic matrices with parameter values separately. In general, this data format limits the variety of potential analyses, because individual matrix elements may be composed of multiple vital rates and this information is lost by storing only the resulting values (_i.e._ the elements of $F$ may be comprised of both probability of reproducing and the per-capita number of propagules produced). To avoid this issue for IPMs, one needs to reconstruct full IPM using the functional form of the kernels and vital rate functions, as well as the associated parameter estimates. One can use tools that associate the symbols with their values to accomplish this task (_e.g._ metaprogramming and _rlang_, Henry & Wickham 2021). _ipmr_ is an _R_ package for users to interactively develop their own IPMs from symbolic model representations and parameter estimates, and perform downstream analyses (Levin et al. 2021). _Rpadrino_ extends this framework to include _reconstructing_ previously published IPMs that are stored in the PADRINO database. 

Here, we introduce _Rpadrino_. _Rpadrino_ provides access to PADRINO, an open access database of IPMs. Specifically, PADRINO houses symbolic representations of IPMs, their parameter values, and associated metadata to aid users in selecting appropriate models. _Rpadrino_ is an R package that enable users to download PADRINO, manage the dataset locally, modify, reconstruct, and analyze IPMs from PADRINO. In the following, we describe how to interact with PADRINO using _Rpadrino_ and discuss future directions for _Rpadrino_ and PADRINO. We also provide two case studies that demonstrate (1) how to use PADRINO and _Rpadrino_ to reconstruct published IPMs, conduct perturbation analyses, compute some life cycle events, and troubleshoot problems, and (2) how to use  _Rpadrino_ and _ipmr_ to combine PADRINO IPMs with user-specified IPMs, and then how to use PADRINO data with other databases with BIEN (Maitner et al. 2017) and COMPADRE as examples. The latter  is intended to demonstrate the potential for _Rpadrino_ in broad, interoperable, macro-ecological applications. Finally, our supplementary materials also contain a detailed overview of how the PADRINO database, along with the associated assumptions and challenges. 


# An introduction to IPMs and PADRINO

First, we provide a brief review of how IPMs are structured. The simplest form of the IPM can be expressed as

$n(z',t+1) = \int_L^U [P(z',z) + F(z',z) + C(z',z)]n(z,t)dz,$

where $n(z',t+1)$ and $n(z,t)$ are the distributions of trait $z$ at time $t+1$ and $t$, $P(z',z)$ is a kernel describing survival and development of existing individuals, $F(z',z)$ is a kernel describing per-capita sexual reproduction, and $C(z',z)$ is a kernel describing per-capita asexual reproduction (_i.e._ clonal reproduction). Each kernel may be comprised of any number of vital rate functions (Ellner, Childs, & Rees 2016). Analytical solutions to the integrals in Eq 1 are not available (Ellner & Rees 2006). Therefore, the integrals are numerically approximated, resulting in a large iteration matrix (typically ranging from $45 \times 45$ to $5000 \times 5000$ in dimension, SC Levin unpublished data), and then some quantities of interest are computed (Ellner, Childs & Rees 2016). 

Before introducing _Rpadrino_, we provide a brief overview of PADRINO. PADRINO is an open-access database of integral projection models. PADRINO defines a syntax to symbolically represent IPMs as text strings, and stores the values of those symbols in a separate table. The syntax used is very similar to the mathematical notation of IPMs, and is largely "language-agnostic" (_i.e._ aims to avoid idiosyncrasies of specific programming languages). For example, a survival/growth kernel with the form $P(z',z) = s(z) * G(z',z)$ would be `P = s * G` in PADRINO's syntax. $G(z',z) = f_G(z'|\mu_g(z), \sigma_G)$ (where $f_G$ denotes a normal probability density function) becomes `G = Norm(mu_g, sd_g)`. This notation should be translatable to many computing languages beyond just _R_ (_e.g._ Python or Julia). Additionally, PADRINO stores extensive metadata to help researchers find IPMs that work for their questions. A more complete description of the database, how IPMs are digitized, and the associated challenges is available in the ESM and the project webpage (https://padrinodb.github.io/Padrino/, Appendix, Tables S1 and S2).

# Rpadrino and ipmr

_Rpadrino_ is an _R_ package that contains functions for downloading the PADRINO database, data querying and management, modifying existing functional forms and parameter values, and reconstructing models. Model reconstruction is powered by the _ipmr_ _R_ package (Levin et al. 2021). While users do not need to know how to use _ipmr_ to use _Rpadrino_, the two package are designed to work with and enhance each other. This means that users can combine IPMs reconstructed with _Rpadrino_ with IPMs of their own constructed with _ipmr_ in a single, coherent analysis (case study 2). Furthermore, users can go from downloading the database to reconstructing IPM objects in as little as 3 function calls. The technical details underlying combining IPM sources are described in _Step 2_ below and graphically in Figure 1.


```{r echo = FALSE, dpi = 250, fig.height = 20, fig.width = 36, fig.cap = "**Figure 1**: An overview of general workflows with *Rpadrino*. The first step is to download and subset PADRINO based on the research question at hand (1a). At this point, if the research question calls for it, users may augment the PADRINO object with data from other sources (1b). The extensive metadata table in PADRINO is designed to make this as straightforward as possible. Once the dataset is prepared, users can also generate automated reports on their data subset. After preparing the dataset, users create `proto_ipm` objects using a single function: `pdb_make_proto_ipm()`. This function translates PADRINO's syntax into *ipmr* code and then uses *ipmr* to generate the `proto_ipm` objects. All of these steps take place internally, and so users do not need to understand PADRINO's syntax or how *ipmr* works to move forward (2). Once `proto_ipm`s are created, users generate IPM objects with a single function, `pdb_make_ipm()`. Examples of how to modify default building settings are in each case study in the supplementary materials and in the function's documentation (3). Finally, *Rpadrino* provides some basic analytical machinery, including deterministic and stochastic growth rates and eigenvectors, the ability to make iteration kernels from sub-kernels, and mean kernels (4). Examples of how to conduct more complex analyses are included in the package vignettes and the case studies in the supplementary materials."}

fig <- readPNG("../figures/figure_1.png")

fig_grob <- rasterGrob(fig, interpolate = TRUE)

plt <- qplot(1:14, 1:14, geom = "blank") +
  annotation_custom(fig_grob,
                    xmin = -Inf,
                    xmax = Inf,
                    ymin = -Inf,
                    ymax = Inf) +
  theme_void()

print(plt)

```

The flexibility of IPMs and their broad application across ecology, evolution, and conservation biology mean that there is no fixed set of steps in a workflow using _Rpadrino_. However, there are generally four steps that a researcher must take when using these _Rpadrino_. The first step is to identify studies of interest (Figure 1, Step 1a), and, optionally, augment PADRINO's metadata with additional information from other sources (_e.g._ environmental data,  GBIF, Figure 1, Step 1b). _Rpadrino_ represents PADRINO objects as a list of `data.frames` (referred to as tables in subsequent text). _Rpadrino_ uses the shared `ipm_id` column across all tables to track information related to each IPM. Therefore, subsetting relies on identifying the correct `ipm_id`'s, and then using those to select the IPMs of interest (Box 1, case study 1 and 2). `data.frames` should be familiar to most _R_ users, and the ability to modify them should readily accommodate the range of further analyses that researchers may be interested in. Users may augment any table with additional information corresponding to, for example, spatial or temporal covariates from other open access databases. Furthermore, _Rpadrino_ provides numerous access functions for metadata that streamline subsetting (Box 1).

The second step in the _Rpadrino_ workflow is to construct a list of `proto_ipm` objects using `pdb_make_proto_ipm()` (Figure 1, Box 1). This function translates PADRINO's syntax into _ipmr_ code, and then builds a `proto_ipm` object for each unique `ipm_id`. For some models, users may choose to create deterministic or stochastic IPMs at this step. _Rpadrino_'s default behavior is to generate deterministic models whenever possible. This behavior encompasses instances where authors generated models with no time or space varying parameters, and where authors included discretely varying environments. The latter can be implemented as deterministic models because all parameter values are known before the IPM is built. IPMs with continuous environmental variation require sampling the environment at each model iteration, usually by sampling from distributions randomly. These are always considered stochastic models. This is also the step where, if needed, users should combine their own `proto_ipm`'s produced by _ipmr_ with the `proto_ipm`'s produced by _Rpadrino_.

The third step in the _Rpadrino_ workflow is creating IPM objects with `pdb_make_ipm()` (Figure 1, Box 1). `pdb_make_ipm()` uses _ipmr_'s `make_ipm()` function to build IPM objects. Users may specify additional options to pass to to `make_ipm()` (_e.g._ normalize the population size to always equal 1, return the vital rate function values as well as the sub-kernels and population state). The various arguments users can modify are described in the _ipmr_ documentation for `make_ipm()`.

The fourth and final step in an _Rpadrino_ workflow is to conduct the analyses of interest (Figure 1, Box 1). _Rpadrino_ provides functions to extract per-capita growth rates, eigenvectors (Caswell 2001, Ellner, Childs & Rees 2016 Ch. 2, demonstrated in Box 1), assess convergence to asymptotic dynamics (Caswell 2001), compute mean kernels for stochastic IPMs (Ellner, Childs & Rees Ch 7), and modify existing IPMs with new parameter values and functional forms. Additionally, the documentation on the Rpadrino website (https://padrinODB.github.io/Rpadrino/index.html) and the Supplementary Materials for this paper contain details on how to conduct more complicated analyses with IPM objects (_e.g._ perturbation analyses (Ellner, Childs & Rees 2016 Ch 4), size at death calculations (Metcalf et al. 2009)). The package documentation and the recent publication describing _ipmr_ also contain code demonstrating analyses on single IPM objects (Levin et al. 2021). These can be extended via the `apply` family of functions.  

```{r eval = FALSE}

# Install and load the CRAN version:
# install.packges("Rpadrino")

library(Rpadrino)

# Step 1 from main text -----
# pdb_download() downloads a copy PADRINO. We can specify a path
# to save the tables using `save = TRUE` and `destination = 'path/to/file/'`.

pdb <- pdb_download(save = FALSE)


# We can use Rpadrino's metadata accessors to get a selection of 'ipm_id's
# that we want to use. For this example, we'll select models for Carpobrotus 
# Geum radiatum. First, extract the 'species_accepted' column. The output of this
# will be named, and the names are the ipm_id associated with each piece of 
# metadata. Thus, we can subset the names of the 'spps' object to get 
# the ipm_id's we need.

spps <- pdb_species_accepted(pdb)
ids  <- names(spps)[spps %in% c("Carpobrotus_spp", "Geum_radiatum")]


# Step 2 from main text -----
# Next, we create a list of proto_ipm's using pdb_make_proto_ipm().
# The second case study demonstrates how to use ipmr and Rpadrino to create
# multiple lists of proto_ipm's and combine them.

my_proto_ipms <- pdb_make_proto_ipm(pdb, ipm_id = ids)

# Step 3 from main text -----
# After creating the proto_ipm list, we can call pdb_make_ipm() to construct
# actual IPM objects.

my_ipms <- pdb_make_ipm(my_proto_ipms)

# Step 4 from main text -----
# After re-building our published IPMs, the next step is to analyze them.
# In this case, we'll just extract the asymptotic population growth rates,
# stable size distribution, and the reproductive values. Note that for the 
# Geum IPMs, there are multiple year-specific values that are returned.

lambdas  <- lambda(my_ipms)
ssds     <- right_ev(my_ipms, iterations = 150, tolerance = 1e-7)
repro_vs <- left_ev(my_ipms , iterations = 150, tolerance = 1e-7)

```

**Box 1**: *An example of a simple analysis workflow using* Rpadrino *. The first step in using *Rpadrino *is to install and load the package. After that, we can use* RPadrino *to download* PADRINO *and, optionally, save it locally on our computer. Once the data are downloaded, we can make use of* Rpadrino*'s metadata accessor functions to quickly select models that meet our criteria (step 1). The concept of the * `ipm_id` *is explained in greater detail in the Appendix of this manuscript. The next step is to use these *`ipm_id`*'s to create a list of *`proto_ipm`*'s using *`pdb_make_proto_ipm()` *(step 2). After this step, we can create actual IPM objects using *`pdb_make_ipm()` *(step 3). Once IPM objects are created, the following steps are according to the demands of the research question. In this case, asymptotic population growth rates, stable size distributions, and reproductive values are extracted (step 4). Note that since the* Geum radiatum *model includes a number of year-specific estimates, multiple values are generated for each quantity we want to extract. The concise representation and reconstruction of models such as this is powered by *ipmr*'s parameter set index notation, which is described in greater detail on the pacakge website (https://levisc8.github.io/ipmr/articles/index-notation.html). However, users do not need to be familiar with this notation unless they wish to modify the IPM in question (see case study 1 for an example of modifying* PADRINO *IPMs with* Rpadrino*).* 

# Challenges

There are numerous challenges associated with reproducing published IPMs. Challenges related to digitizing and storing IPMs are discussed in Appendix 1. Important challenges remain in the reconstruction of IPMs. Semi- or non-parametric models may be used to generate IPMs whose functional form is not known _a priori_. We have not yet developed a general syntax for representing these models in PADRINO, though work is ongoing. Additionally, _ipmr_ is not yet able to handle two-sex models (_e.g._ Stubberud et al. 2019), time-lagged models (_e.g._ Rose et al. 2005), or periodic models (_e.g._ Letcher et al. 2014). These types of IPMs do not yet represent a substantial portion of the literature. Nonetheless, it is our intention to continue developing functionality to accommodate them in future releases of _Rpadrino_, _ipmr_, and PADRINO. 

# Opportunities and Future Directions

_Rpadrino_ presents unique opportunities for synthesis in both theoretical and applied contexts. The expanded range of phylogenetic and geographical coverage can be used in conjunction with other demographic databases (_e.g._ COM(P)ADRE (Salguero-Gómez et al. 2015, Salguero-Gómez et al. 2016), popler (Compagnoni et al. 2019), DatLife (DatLife 2021)) to power larger scale syntheses than were possible before (_e.g._ Compagnoni et al. 2021b). For example, one could use IPMs from PADRINO and matrix population models from COMPADRE and COMADRE to create life tables (Jones et al. 2021), which could then be combined with life tables from DATLife for further analysis (_e.g._ Jones et al. 2014). The intermediate life table conversion steps may not be necessary, as many of the same life history traits and population level parameters may be calculated from all of these models (Caswell 2001, Ellner, Childs & Rees 2016). Furthermore, recent publications combine biotic and abiotic interactions into demographic models providing a robust theoretical toolbox for exploring species responses to environmental drivers such as climate change (_e.g._ Simmonds et al. 2020, Abrego et al. 2021). _Rpadrino_ also provides functionality to modify parameter values and functional forms of the IPMs it stores, giving theoreticians a wide array of realistic life histories to experiment with. These features will enable researchers to carry out more detailed and comprehensive analyses at various spatial, temporal, and phylogenetic scales. The examples given here are far from an exhaustive list, but hopefully demonstrates the potential for this new tool in demography, ecology, and evolutionary biology. 


```{r echo = FALSE}

cap <- "**Table 1**: Taxonomic representation of IPMs accesible via Rpadrino. These numbers represent the number of models that are error checked and accurately reproduce the published IPM (see 'Data Validation' in the Appendix for more details). Models that are partially entered or still contain errors are not considered here. We are in the process of correcting them and/or retrieving additional information from the authors. See Appendix for details."
  
kable(tot,
      col.names = c("Kingdom", 
                    "# of Unique ipm_id's", 
                    "# of Unique Species",
                    "# of Publications"),
      caption = cap)

```

## Citations

1. Easterling, M.R., Ellner, S.P., & Dixon, P.M. (2000). Size specific sensitivity: applying a new structured population model. Ecology 81(3): 694-708.

2. Ellner, S.P. & Rees, M. (2006). Integral Projection Models for species with complex demography. The American Naturalist 167(3): 410-428.

4. DATLife – The Demography Across the Tree of Life – database. Max-Planck Institute for Demographic Research (Germany). Available at www.datlife.org. (2021)

5. Salguero‐Gómez, R., Jones, O.R., Archer, C.R., Bein, C., de Buhr, H., Farack, C., Gottschalk, F., Hartmann, A., Henning, A., Hoppe, G., Roemer, G., Ruoff, T., Sommer, V., Wille, J. Voigt, J., Zeh, S., Vieregg, D., Buckley, Y.M., Che-Castaldo, J., Hodgson, D., et al. (2016) COMADRE: a global database of animal demography. Journal of Animal Ecology 85: 371-384. https://doi.org/10.1111/1365-2656.12482

6. Salguero-Gómez, R, Jones, O.R., Archer, C.A., Buckley, Y.M., Che-Castaldo, J., Caswell, C., Hodgson, D., Scheuerlein, A., Conde, D.A., Brinks, E., de Buhr, H., Farack, C., Gottschalk, F., Hartmann, A., Henning, A., Hoppe, G., Roemer, G., Runge, J., Ruoff, T., et al. (2014) The COMPADRE Plant Matrix Database: an online repository for plant population dynamics. Journal of Ecology 103: 202-218. https://doi.org/10.1111/1365-2745.12334

7. Compagnoni, A., Pardini, E., & Knight, T.M. (2021a). Increasing temperature threatens an already endangered coastal plant species. Ecosphere 12(3): e03454. https://doi.org/10.1002/ecs2.3454

8. Compagnoni, A., Levin, S.C., Childs, D.Z., Harpole, S., Paniw, M., Roemer, G., Burns, J.H., Che-Castaldo, J., Rueger, N., Kunstler, G., Bennett, J.M., Archer, C.R., Jones, O.R., Salguero-Gómez, R., & Knight, T.M. (2021b). Herbaceous perennial plants with short generation time have stronger responses to climate anomalies than those with longer generation time. Nature Communications 12: 1824. https://doi.org/10.1038/s41467-021-21977-9

9. Levin, S.C., Childs, D.Z., Compagnoni, A., Evers, S., Knight, T.M., & Salguero-Gómez, R. (2021) ipmr: Flexible implementation of Integral Projections Models in R. Methods in Ecology and Evolution. https://doi.org/10.1111/2041-210X.13683

10. Jongejans, E., Shea, K., Skarpaas, O., Kelly, D., & Ellner, S.P. (2011). Importance of individual and environmental variation for invasive species spread: a spatial integral projection model. Ecology 92(1): 86-97. https://doi.org/10.1890/09-2226.1

12. Erickson, R.A., Eager, E.A., Brey, M.B., Hansen, M.J., & Kocovsky, P.M. (2017). An integral projection model with YY-males and application to evaluating grass carp control. Ecological Modelling 361: 14-25. https://doi.org/10.1016/j.ecolmodel.2017.07.030

13. Childs, D.Z., Rees, M., Rose, K.E., Grubb, P.J., & Ellner, S.P. (2004). Evolution of size-dependent flowering in a variable environment: construction and analysis of a stochastic integral projection model. Proceedings of the Royal Society B 271(1547): 425-434. https://doi.org/10.1098/rpsb.2003.2597

14. Ellner, S.P., Childs, D.Z., Rees, M. (2016) Data-driven modelling of structured populations: a practical guide to the integral projection model. Basel, Switzerland: Springer International Publishing AG

15. Leslie, P.H. (1945). The use of matrices in certain population mathematics. Biometrika 33(3): 183-212.

16. Caswell, H. (2001) Matrix population models: construction, analysis, and interpretation, 2nd edn. Sunderland, MA: Sinauer Associates Inc

17. Adler, P.B., Ellner, S.P. & Levine, J.M. (2010). Coexistance of perennial plants: an embarassment of niches. Ecology Letters 13: 1019-1029. https://doi.org/10.1111/j.1461-0248.2010.01496.x

18. Abrego, N., Roslin, T., Huotari, T., Ji, Y., Schmidt, N.M., Wang, J., Yu, D.W., & Ovaskainen, O. (2021) Accounting for species interactions is necessary for predicting how arctic arthropod communities respond to climate change. Ecography. https://doi.org/10.1111/ecog.05547 

19. Henry, L., & Wickham, H. (2021). rlang: Functions for Base Types and Core R and 'Tidyverse' Features. R package version 0.4.11. https://CRAN.R-project.org/package=rlang

20. Crone, E.E., Menges, E.S., Ellis, M.M., Bell, T., Bierzychudek, P, Ehrlen, J. et al. (2011) How do ecologists use matrix population models? Ecology Letters 14(1): 1-8. DOI: https://doi.org/10.1111/j.1461-0248.2010.01540.x

21. Doak, D. & Morris W.F. (2002). Quantitative Conservation Biology: Theory and Practice of Population Viability Analysis. Oxford, UK: Oxford University Press

22. Salguero-Gómez, R. & Gamelon M. (2021). Demographic Methods Across the Tree of Life. Oxford, UK: Oxford University Press

23. Coulson, T., MacNulty, D.R., Stahler, D.R., von Holdt, B., Wayne, R.K., & Smith, D.W. (2011). Modeling effects of environmental change on wolf population dynamics, trait evolution, and life history. Science 334(6060): 1275-1278. https://doi.org/10.1126/science.1209441

24. Stott, I., et al. (2011). A framework for studying transient dynamics of population projection matrix models. Ecology Letters 14: 959-970. DOI: https://doi.org/10.1111/j.1461-0248.2011.01659.x

25. Clark, J.S. (2003) Uncertainty and variability in demography and population growth: A hierarchical approach. Ecology 84(6): 1370-1381. 

26. Letcher, B.H., Schueller, P., Bassar, R.D., Nislow, K.H., Coombs, J.A.,Sakrejda, K. _et al._ (2014). Robust estimates of environmental effects on population vital rates: an integrated capture-recapture model of seasonal brook trout growth, survival and movement in a stream network. Journal of Animal Ecology 84(2): 337-352. https://doi.org/10.1111/1365-2656.12308

27. Stubbered, M.W., Vindenes, Y., Vollestad, L.A, Winfield, I.J., Stenseth, N.C., & Langangen, O. (2019). Effects of size- and sex-selective harvesting: an integral projection model approach. Ecology and Evolution 9: 12556-12570. https://doi.org/10.1002/ece3.5719

28.  Maitner, B., Boyle B., Casler N., Condit R., Donoghue J., Duran S.M.,  _et al._ (2017) The bien r package: A tool to access the Botanical Information and Ecology Network (BIEN) database.  Methods in Ecology and Evolution 9(2): 373-379. https://doi.org/10.1111/2041-210X.12861

29. Simmonds E.G., Cole, E.F., Sheldon, B.C., & Coulson, T. (2020). Phenological asynchrony: a ticking time-bomb for seemingly stable populations? Ecology Letters 23(12): 1766-1775. https://doi.org/10.1111/ele.13603

30. Kousta, S., Pastrana, E., & Swaminathan, S. (2019). Three approaches to support reproducible research. Science Editor 42(3): 77-82.

31. Jones, O.R., Scheuerlein, A., Salguero-Gómez, R., Camarda, C.G., Schaible, R., Casper, B.B., _et al._ (2014). The diversity of ageing across the tree of life. Nature 505: 169-173. https://doi.org/10.1038/nature12789

32. Jones, O.R., Barks, P., Stott, I., James, T.D., Levin, S.C., Petry, W.K. _et al._ (2021). Rcompadre and Rage - two R packages to facilitate the use of the COMPADRE and COMADRE databases and calculation of life history traits from matrix population models. bioRxiv. https://doi.org/10.1101/2021.04.26.441330

## Acknowledgments

Funding: R.S-G. was supported by a NERC Independent Research Fellowship (NE/M018458/1). SCL, AC, SE, and TMK were funded by the Alexander von Humboldt Foundation in the framework of the Alexander von Humboldt Professorship of TM Knight endowed by the German Federal Ministry of Education and Research. 

Author Contributions: SCL designed _ipmr_ and _Rpadrino_ with contributions from all authors, and SCL implemented the packages. SCL wrote the first draft of the manuscript and all authors provided comments.
